---
title: "Example R Markdown Analysis"
author: "Jasmine Hughes"
date: "9/17/2021"
output:
  html_document:
    df_print: paged
  pdf_document:
    template: template.tex
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE,message=FALSE)
```

# Load Data 

In this part of my analysis, I can write plain text.

It's useful for introducing the reader to my data set, like explaining how the 
data was collected, or the motivation behind my project.

```{r load_libraries}
library(ggplot2)
library(dplyr)
```

In between text, I can intersperse bits of code in what are called "chunks".

Chunks start and end with three backticks (not apostrophes!). Then, curly 
brackets ({}) are used to describe settings for the chunk.

The first piece of information to share is the language. We are going to use R, 
so it should include an `r`:

```{r}
my_data_set <- read.csv("../data/iris.csv")
# note the file path here starts with ../
# ".." means "go up one folder in the path"
# Rmarkdown notebooks execute file paths relative to where the Rmd file is saved
```

# Create a graph

Chunks can make coding easier, because you can work interactively:

```{r create_a_graph,fig.width=4,fig.height=3}

ggplot(my_data_set) +
  aes(x = Sepal.Length, y = Sepal.Width, color = Species) +
  geom_point() +
  theme_minimal() +
  xlab("Length") +
  ylab("Width") +
  scale_color_manual(values = c("magenta4", "orangered2", "turquoise4"))

```

# Save the output of our code

```{r save_graph, echo=FALSE}
my_graph <- ggplot(my_data_set) +
  aes(x = Sepal.Length, y = Sepal.Width, color = Species) +
  geom_point() +
  theme_minimal() +
  ggtitle("Sepal length and width for different iris species") +
  xlab("Length") +
  ylab("Width") +
  scale_color_manual(values = c("magenta4", "orangered2", "turquoise4"))

ggsave("iris_graph_rmd.png")

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

# Add whatever you need to make your code useful!

You could write (and I have written!) publication-ready papers in R markdown!

## Equations
$mean = \frac{\sum{x}}{n}$

## Citations

---
references:
- id: Asimov1959
  title: Enzymes and Metaphor
  author:
  - family: Asimov
    given: Isaac
  container-title: Journal of Chemical Education
  volume: 36
  URL: 'https://doi.org/10.1021/ed036p535'
  DOI: 10.1021/ed036p535
  issue: 11
  publisher: ACS Publications
  page: 261-263
  type: article-journal
  issued:
    year: 1959
    month: 11
---

Is it even science if you don't cite your sources? There are many ways to refer 
to your favorite papers [@Asimov1959], but here's an example down below.



# References
