---
title: "2) Setting up your R work space"
author: "Jasmine Hughes"
date: "9/17/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup2, eval=TRUE}
if (!require("ggplot2")) {
  install.packages("ggplot2")
}
if (!require("dplyr")) {
  install.packages("dplyr")
}
```

# RStudio

* RStudio is an integrated development environment (IDE)
* Widely used way of writing R code and interacting with data.
* RStudio is to R and Microsoft Word is to English. 

Let's take a little tour:

1. Source: R code files:
    * ".R" files: scripts
    * ".Rmd" files: "R Markdown", interactive notebooks that can be saved as 
       LaTeX/pdf or HTML documents. Handy for exploring data, making graphs, 
       and keeping track of why you are making particular decisions.
    * I will mostly be using R Markdown notebooks in this course, but you can
      use either R or Rmd files!
  
2. Console
    * Another way to code interactively!
    * Also has a terminal (the same as terminal/command line)

3. Files/Plots/Help
    * Useful references for while doing data analysis and writing code

4. Environment
    * variables that you've saved that are available for you to use.