---
title: '3) The Basics'
author: "Jasmine Hughes"
date: "9/17/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R as a calculator

```{r computing,echo=TRUE,results='markup'}
2 + 2  # add numbers
2 * 3  # multiply numbers
3 ^ 4  # powers
3 ** 4 # powers
2 * pi # multiply by a constant
(3 + 4) * 2 # order of operations
```

```{r computing2,echo=TRUE,results='markup'}
sqrt(4^4) # square root
23 %% 2   # remainder
log(10)   # logarithms (what is the default base?)
log(100, base = 10)  # you can pass other arguments to change the base
?log
exp(1)

# scientific notation works!
5000000000 * 1000
5e9 * 1e3
```

Think of a mathematical operation you need - can you guess how to do it in R?

# Assigning values to R objects

So far we have been mostly coding "interactively" - running a single command and 
looking at the answer - or we have been saving objects (like graphs saved as 
pictures) to our "working directory".

But what if we want to use the same object for later calculations?

A key action in R is to store values in the form of R objects, and to examine 
the value of R objects. (Add notes to help remind yourself what's going on!)

```{r assigning_values}
# (variable name) <- (value, or code to create a value/object)
val <- 5

# (variable name) = (value, or code to create a value/object)
val = 6 # equal sign *can* be used for assignment but is not conventional

# In both the console and in R chunks, 
# running a variable name will print its value out
val
print(val)

Val <- 7 # case-sensitive!
print(c(val, Val))

val <- 9 # can be overwritten
print(c(val, Val))



```

You can see these values change in the "Environment" tab in Rstudio.

We can work with (and store) sequences and repetitions
```{r sequences}
mySeq <- 1:6
mySeq

years <- seq(1952, 2007, by = 5)
years
length(years)

country <- rep("Afghanistan", 12)
country 

## This is a comment: here is an example of non-numeric data
countries <- c("Afghanistan", "India", "Pakistan")
countries
```

Try calculating and saving values as variables in the chunk below:

```{r try_it_yourself}

```


# Vectors in R

The most basic form of an R object is a vector. In fact, individual (scalar) 
values are vectors of length one. The variables we created earlier (val, Val, 
mySeq, years) are all vectors!

We can concatenate values into a vector with `c()`.

## 1) Numeric vectors
Also known as "double precision floating point"

```{r types_of_vectors1}
nums <- c(1.1, 3, -5.7)
myvals <- c(val, Val)
devs <- rnorm(5)
devs
```

## 2) Integer vectors
Less common than numeric vectors, but it's good to know they exist since
sometimes you'll see errors related to the difference between numeric and 
integer vectors
```{r types_of_vectors2}

ints <- c(1L, 5L, -3L) # force storage as integer not decimal number
## R Trivia: 'L' stands for 'long integer' (historical)
notints <- c(1, 5, -3)
identical(ints, notints)
```
## 3) character vector
Most common way of storing non-numeric data

```{r types_of_vectors3}
chars <- c("pine", "apple")
paste(chars, collapse = '')
paste("character", "vector", sep = " ")

chars <- c('hi', 'hallo', "mother's", 'father\'s', 
   "She said, 'hi'", "He said, \"hi\"" )  
# the backslash is an "escape character"
# make sure opening and closing characters match!
chars
cat(chars[5])
cat(chars[6])

```
## 4) logical vector (or boolean vector)
```{r types_of_vectors4}

bools <- c(TRUE, FALSE, TRUE)
bools2 <- c(T, F, T) # less preferred (less legible)
bools
identical(bools, bools2)
val == 8
val <- 8 
val = 9
val < 9 

val != 10
```

```{r types_of_vectors5}
## combination
combi <- c("pizza", 9, TRUE)
combi

# How do I know what datatype something is?
typeof(combi)
```

# Working with indices and subsets

```{r indexing}
vals <- seq(2, 12, by = 2)
vals
vals[4]
```

Can you guess what these will do before running the code??
```{r indexing2}
vals[3:5]
vals[c(1, 3)]
vals[-c(1, 3)]
which_indices <- c(rep(TRUE, 3), rep(FALSE, 2), TRUE)
vals[which_indices]
```

# Comparing Values
```{r comparisons}

vec1 <- c(1, 2, 3, 4, 5)
vec2 <- c(7, -4, 3, 4, 5.1)
vec2 < vec1
vec1 <= vec2

vec1 == vec2
vec1 != vec2
## careful: 
vec1 = vec2
identical(vec1, vec2)

## using 'or'
vec1 >= 3 | vec1 == exp(0) | vec1 < -7
## using 'and'
vec1 >= 3 & vec1 %% 2 == 0 

10 %% 2
11 %% 2
```

# Reading Data From A File

```{r load_data}

#######################################################################
## IMPORTANT: read in the Gapminder dataset from disk;
## first make sure your working directory is the 'workbooks' directory:
getwd()
## if the result is not the 'workbooks' subdirectory of the bootcamp
## directory, set the working directly along the lines of this:
##
## setwd('/Users/jasminehughes/Documents/Intro_to_R/RforPharmacometricians2021')
##
## replace '/Users/jasminehughes/Documents/Intro_to_R/' with whatever directory 
## you put the bootcamp materials in; e.g. on Windows it might be something like
## 'C:\\Users\\sarah\\R_for_pharmacometricians'
##
## You can also try navigate to this directory through the "Files" window,
## then click The "More (cogwheel)" > Set as working directory
##
## If you've done that correctly, then the next command reads
## in the dataset from the 'data' directory. In the next
## command R finds that directory relative to the current
## working directory.

gap <- read.csv(
  file.path("..", "data", "gapminder-FiveYearData.csv"),
  stringsAsFactors = FALSE
)



#######################################################################
```
Our data was imported as a dataframe. Dataframes are tabular: defined by rows and columns. When doing statistics & data analysis in R, you'll mostly be working with dataframes.

```{r looking_at_your_data}
## Look at part of the data
head(gap) # first few lines)
tail(gap) # last few lines
View(gap) # the whole shebang (avoid with very large datasets)
dim(gap) # dimensions
summary(gap) # basic info about each column
```

# R is a functional language

* Operations are carried out with functions. Functions take objects as inputs 
  and return objects as outputs. 
* An analysis can be considered a pipeline of function calls, with output from a 
  function used later in a subsequent operation as input to another function.
* Functions themselves are objects: 

```{r functions1}
median
class(median)
median(c(6, 7, 18, 9.2))
median(gap[["lifeExp"]])
```

Functions generally take arguments, some of which are often optional:
```{r functions2}
maxYear <- max(gap$year)
lifeExp <- gap$lifeExp[gap$year == maxYear]
median(lifeExp)
# We just calculated the median life expectancy in 2007!
```

* We can embed function calls: 

```{r, hist, fig.cap = ""}
random_data <- rnorm(1000)
hist(random_data)

hist(rnorm(1000))
```

# Getting help about a function

To get information about a function you know exists, use `help` or `?`, 
e.g., `?lm`. For information on a general topic, use `apropos`.

```{r eval = FALSE}
help(lm)
?lm

?log

apropos("mean")

```

# Basic kinds of R objects

Vectors are not the only kinds of R objects.

### Vectors 
We saw these earlier!

Vectors of various types (numeric (i.e., decimal/floating point/double), 
integer, boolean, character), all items must be of the same type

### Data frames
This is what R does best!

Collections of columns of potentially different types

If you're at home in excel spreadsheets, this might feel most intuitive.

```{r things_to_do_with_dfs}
head(gap, 10)
dim(gap)
nrow(gap)
names(gap)
class(gap)
is.matrix(gap)
class(gap$year)
class(gap$lifeExp)
class(gap$country)

```


# Converting between different types of objects

You can use the `as` family of functions.

```{r converting_between_types}
ints <- 1:10
ints[4]
as.character(ints)
as.numeric(c('3.7', '4.8'))
```

Be careful: R tries to be helpful and convert between types/classes when it 
thinks it's a good idea. Sometimes it is overly optimistic. 

```{r}
indices <- c(1.7, 2.3, 4.6)
ints[indices]
ints[0.999999999]
```


# Be Lazy

If you're starting to type something you've typed before, or the long name of an 
R object or function, STOP!  You likely don't need to type all of that.

- Tab completion
- Command history 
    * up/down arrows
    * Ctrl-{up arrow} or Command-{up arrow}
- RStudio: select a line or block for execution
- Write re-usable code! Put your code in a file and use `source()`. 
  Example: `source(file.path('..', 'example_scripts', 'Example_Rscript.R'))`. 

Or, create functions you can use within an analysis.

# Common issue: incomplete code

Sometimes you write some code that isn't a "complete sentence" and R gets
confused. 
```{r eval=FALSE}
(9 + 2
```
The R console will show a little + sign, indicating it is expecting more.

Press escape to try again, or complete the "sentence" by adding the missing 
bracket.

# Breakout

In general, your answers to any questions should involve writing code to 
manipulate objects. For example, if I ask you to find the maximum flight delay, 
do not scan through all the values and find it by eye. Use R to do the 
calculations and print results.

Some of these questions will require you to search for the right function.

**Hint**: use the function `apropos("key word")` to find something to use.


1) The variable y contains all the life expectancy data in the "gap" data frame 
we loaded earlier. Create a variable called 'x' that contains the mean life 
expectancy. Then, create a vector of length 2 that contains the minimum and 
maximum life expectancy. 
```{r}
y <- gap[["lifeExp"]]
y <- gap$gdpPercap
#This is where I write my answers!

```

2) Use functions in R to round 'x' to two decimal places and to two 
significant digits.

```{r}

```

3) Create a vector of GDP per capita in units of Euros rather than dollars.

```{r}


```

4) Create a boolean (TRUE/FALSE) vector indicating whether total country GDP is greater than 1 trillion dollars. When entering 1 trillion, use R's scientific notation.

```{r}

```
