---
title: 'Module 4: Managing your R Environment'
author: "Jasmine Hughes"
date: "9/17/2021"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

# Managing and saving the workspace

R has functions for learning about the collection of objects in your workspace. Some of this is built in to RStudio.

```{r}
x <- rnorm(5)
y <- c(5L, 2L, 7L)
z <- list(a = 3, b = c('sam', 'yang'))
ls()  # search the user workspace (global environment)
rm(x)    # delete a variable
ls()
ls.str() # list and describe variables
```

Finally we can save the objects in our R session:
```{r}
ls()
save.image('module4.Rda')
rm(list = ls())
ls()
load('module4.Rda') 
ls()
```

Although you should try to write R scripts that are reproducible, it can still 
be convenient to save all your R objects so you can easily pick up where you 
left off last time.

```{r, include=FALSE}
# load gapminder data back in for use later
gap <- read.csv(
  file.path('..', 'data', 'gapminder-FiveYearData.csv'), 
  stringsAsFactors = FALSE
)
```


# Packages (R's killer app)

Let's check out the [packages on CRAN](https://cran.r-project.org/web/packages/) 

Essentially any well-established and many not-so-established statistical methods 
and other functionality is available in a package. 

If you want to sound like an R expert, make sure to call them *packages* and 
not *libraries*. A *library* is the location in the directory structure where 
the packages are installed/stored.

### Using packages

Two steps:

1. Install the package on your machine
2. Load the package

To install a package, in RStudio, just do `Packages->Install Packages`.

From the command line, you generally will just do
```{r eval = FALSE}
install.packages('dplyr') 
```
That should work without specifying the repository from which to download the 
package (though sometimes you will be given a menu of repositories from which to 
select) but sometimes you'll get errors indicating a package is not available 
for your version of R if you don't include an explicit repository. 

If you're on a network and are not the administrator of the machine, you may 
need to explicitly tell R to install it in a directory you are able to write in:
```{r eval = FALSE}
install.packages('dplyr', lib = file.path('~', 'R'))
```

Now would be a good point to install the various packages we need for the bootcamp, which can be done easily with the following command:

```{r eval=FALSE}
install.packages(
  c(
    'tidyverse',    #  also sorts of data science tools!
    'devtools',     #  making R packages and other more advanced tools
    'ggplot2',      # graphing!
    'gridExtra',    # more graphing!
    'haven',        # reading/writing specialized data types
    'knitr',        # Rmarkdown tools
    'scales',       # customized graph axes
    'reprex',       # reproducible examples
    'xlsx'          # working with excel
  )
)
```

Note that packages often are dependent on other packages so these dependencies 
may be installed and loaded automatically.

You can also install directly from a package zip/tarball rather than from CRAN 
by giving a filename instead of a package name.

### General information about a package

You can use syntax as follows to get a list of the objects in a package and a 
brief description: `library(help = packageName)`. 

On CRAN there often *vignettes* that are an overview and describe usage of a 
package if you click on a specific package. The *reference manual* is just a 
single document with the help files for all of the objects/functions in a 
package, so may be helpful but often it's hard to get the big picture view from 
that.

# More on packages

### The search path

To see the packages that are loaded and the order in which packages are searched 
for functions/objects: `search()`.

```{r}
search()
original_search <- search()
library(lubridate) # a tidyverse package
search_after_library_load <- search()
setdiff(search_after_library_load, original_search)
```

To see what *libraries* (i.e., directory locations) R is retrieving packages 
from: `.libPaths()`.

And to see where R is getting specific packages, `searchpaths()`.

### Package namespaces

Namespaces are way to keep all the names for objects in a package together in a 
coherent way and allow R to look for objects in a principled way.

A few useful things to know:

```{r}
library(dplyr)
# You can reference functions within a package using two colons:
stats::t.test(rnorm(15, 6), rnorm(20, 7))

# This is useful when two functions have the same name:
ls('package:stats')[1:20]
lm <- function(i) {
   print(i)
}
lm(7) 
lm(gap$lifeExp ~ gap$gdpPercap)
stats::lm(gap$lifeExp ~ gap$gdpPercap)
rm(lm)


```

Can you explain what is going on? Consider the results of `search()`.


### Looking inside a package

Packages are available as "Package source", namely the raw code and help files, 
and "binaries", where stuff is packaged up for R to use efficiently. 

To look at the raw R code (and possibly C/C++/Fortran code included in some 
packages), download and unzip the package source tarball (.tar file), or check 
out the code on github.


### Creating your own R package

R is do-it-yourself - you can write your own package! At its most basic this is 
just some R scripts that are packaged together in a convenient format. And if 
giving it to someone else, it's best to have some documentation in the form of 
function help files. 

Why make a package?

* It's an easy way to share code with collaborators
* It's a good way to create self-contained code for code you commonly use yourself
* It's how you can share your code and methods with the outside world
* It helps make your work reproducible
* It forces you to be more formal about your coding, which will improve your code

See the *devtools* package and `package.skeleton()` for some useful tools to 
help you create a package. And there are lots of tips/tutorials online, in 
particular [Hadley Wickham's R packages book](https://r-pkgs.had.co.nz/).

### When should I create my own package?

* If you use the same code more than once: make a function
* If you use the same function in more than one project or assignment, 
  make a package.
* You don't need to be an expert to make a package!

# The working directory

To read and write from R, you need to have a firm grasp of where in the 
computer's filesystem you are reading and writing from. 

```{r, eval=FALSE}
getwd()  # what directory will R look in?
# Linux/Mac specific
setwd('~/Documents') # change the working directory
setwd('/Users/jasminehughes/Documents') # absolute path
getwd()
setwd('Intro_to_R/R_for_pharmacometricians/Modules') # relative path


# Windows - use either \\ or / to indicate directories
# setwd('C:\\Users\\Your_username\\Desktop\\R_for_pharmacometricians')
# setwd('..\\R_for_pharmacometricians')

# platform-agnostic
setwd(
  file.path(
    '/', 'Users', 'jasminehughes', 'Documents', 
    'Intro_to_R', 'RforPharmacometricians2020'
  )
) # absolute path
getwd()
setwd(file.path('data')) # relative path

# check what is in a folder
list.files()
```
Many errors and much confusion result from you and R not being on the same page in terms of where in the directory structure you are.

# Reading text files into R

The workhorse for reading into a data frame is `read.table()`, which allows any 
separator (CSV, tab-delimited, etc.). `read.csv()` is a special case of 
`read.table()` for CSV files.

Here's a simple example where R is able to read the data in using the default 
arguments to `read.csv()`.

```{r, cache=TRUE}
medals <- read.csv(file.path('..', 'data', 'all_medalists.csv'))
head(medals)
medals
```

It's good to first look at your data in plain text format outside of R and then 
to check it after you've read it into R.

Careful about handling [genomic information in excel](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-016-1044-7)!

# Reading 'foreign' format data

Here's an example of reading data produced by another statistical package 
(Stata) with `read_dta()`. 

```{r}
if(!require('haven')) install.packages('haven')
library(haven)
vote <- read_dta(file.path('..', 'data', '2004_labeled_processed_race.dta'))
head(vote)
```

There are a number of other formats that we can handle for either reading or 
writing. Let's see `library(help = haven)`.

R can also read in (and write out) Excel files, netCDF files, HDF5 files, etc., 
in many cases through add-on packages from CRAN. 

Best practices for data-sharing: avoid using proprietary formats like Excel. 
TXT and CSV files are much easier to read across platforms and programming 
languages.

# Writing data out from R

Here you have a number of options. 

1) You can write out R objects to an R Data file, as we've seen using `saveRDS()`.
2) You can use `write.csv()` and `write.table()` to write data frames to flat 
   text files with delimiters such as comma and tab.
3) You can use `write()` to write out matrices in a simple flat text format.
4) You can use `cat()` to write to a file, while controlling the formatting.
5) You can write out in the various file formats mentioned on the previous slide


# Breakout

### Basics

1) Make sure you are able to install packages from CRAN. E.g., try to install 
*praise*. What does the code below do?

```{r }
praise::praise()
```

2) Figure out what your current working directory is.

```{r }
getwd()
```

### Using the ideas

3) Put the *data/all_medalists.csv* file in some other directory on your 
computer, such as *Downloads*. Use `setwd()` to set your working directory to be 
that directory. Read the file in using `read.csv()`.  Now use `setwd()` to point 
to a different directory such as *Desktop*. Write the data frame out to a file 
without any row names and without quotes on the character strings.

```{r }

write.csv(file = "iris.csv", x = iris)
```

4) (Optional) If you have a .csv file or a Microsoft Excel file with data that 
you've used for an unrelated project or assignment, try open it now, just to see 
if you can!

```{r}
# This function might be useful:
xlsx::read.xlsx()
```


# Other ways to read data into R

The *read.table()* family of functions just skims the surface of things...

1) You can also read in a file as vector of characters, one character string per 
   line of the file with *readLines()*, and then post-process it. 
2) You can read fixed width format (constant number of characters per field) 
   with *read.fwf()*.
3) *read_csv()* (and *read_lines()*, *read_fwf()*, etc.) in the *readr* package 
   is a faster, more helpful drop-in replacement for *read.csv()* that plays well 
   with *dplyr* (see Workbook 5).
4) the *data.table* package is great for reading and manipulating large datasets 
   (orders of gigabytes or 10s of gigabytes).

# Version control

### Overview

At a basic level, a simple principle is to have version numbers for all your 
work: code, datasets, manuscripts. Whenever you make a change to a dataset, 
increment the version number. For code and manuscripts, increment when you make 
substantial changes or have obvious breakpoints in your workflow. 

However, this is a hassle to do manually. Instead of manually trying to keep 
track of what changes you've made to code, data, documents, you use software to 
help you manage the process. This has several benefits:

* easily allowing you to go back to earlier versions
* allowing you to have multiple versions you can switch between
* allowing you to share work easily without worrying about conflicts
* providing built-in backup


### Git and Github

Git is a popular tool for version control. Git is based around the notion of a 
repository, which is basically a version-controlled project directory. 
Many people use it with the Github or Bitbucket online hosting services for 
repositories.

In the introductory material, we've already seen how to get a copy of a Github 
repository on your local machine. 

"Track changes" but for code.

# Getting R help online

### Mailing lists

There are several mailing lists that have lots of useful postings. In general if 
you have an error, others have already posted about it.

- [Stack Overflow](https://stackoverflow.com) (R stuff will be tagged with [R])
- Simple Google searches 
    - You may want to include "in R", with the quotes in the search
- Reddit: /r/rprogramming

If you are searching you often want to search for a specific error message. 
Remember to use double quotes around your error message so it is not broken into 
individual words by the search engine. 

### Posting your own questions

The main rule of thumb is to do your homework first to make sure the answer is 
not already available on the mailing list or in other documentation. Some of the 
folks who respond to  questions are not the friendliest so it helps to have a 
thick skin, even if you have done your homework. On the plus side, they are very 
knowledgeable and include the world's foremost R experts/developers.

Here are some guidelines when posting to one of the R mailing lists [https://www.r-project.org/posting-guide.html](https://www.r-project.org/posting-guide.html)

`sessionInfo()` is a function that will give information about your R version, 
OS, etc., that you can include in your posting.

You also want to include a short, focused, [reproducible](https://adv-r.had.co.nz/Reproducibility.html) 
example of your problem that others can run. 

```{r}
library(lubridate)
example_function <- function(y) {
  hours(y + 1)
}
example_function(7)
# copy the above to your clipboard, then run the line below:
reprex::reprex()

# how do we fix the error?
```
