---
title: '11) Programming in R'
author: "Jasmine Hughes"
date: "9/17/2021"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
library(dplyr)
gap <- read.csv(
  file.path("..", "data", "gapminder-FiveYearData.csv"), 
  stringsAsFactors = FALSE
) 
```

# Overview

So far, most of the tasks we've carried out in R have been basic operations, or 
have leveraged the `tidyverse` and other packages for graphing or working with 
data frames. However, sometimes you might need to perform somewhat more 
complicated tasks. In this notebook, we will look at some of the basic building 
blocks that you can put together to accomplish these tasks.

These are some of the key pieces missing from SAS, Stata, SPSS and other tabular 
data analysis software.


# Loops
```{r example_loop}
for (iter in 1:10){
  print(paste("the value of iter is", iter))
}

```

In many languages, looping (for loops, while loops, etc.) is one of the main 
constructs used to carry out computation. Loops are not emphasized as much in R, 
both because they can be slow and because other syntax is often cleaner.

But there are lots of times when using a loop does make sense.

Most of you are probably familiar at least with the basic idea of iterating 
through a series of steps. A for loop iterates through a pre-determined number 
of iterations, while a while loop iterates until some condition is met. For 
loops are more common in R, but while loops can be handy particularly for things 
like optimization.

## For loops

```{r, for-example}
out <- list()
years <- unique(gap$year)
length(out) <- length(years)
names(out) <- years

for (yrIdx in seq_along(years)) {
     # equivalently: for(i in 1:length(DestSubset))
     # n.b., seq_along(x) is better than 1:length(x), since it handles the case
     # where the length of an object is 0 or NULL more robustly.
     
  # select the year of this iteration 
  sub <- filter(gap, year == years[yrIdx])
   
     
  if (sum(!is.na(sub$lifeExp)) && length(sub$lifeExp) > 1) {
  # as long as data isn't missing...
  # fit regression
    out[[yrIdx]] <- lm(lifeExp ~ log(gdpPercap), data = sub)
  } else {
    out[[yrIdx]] <- NA
  }
}
out[['2007']]
summary(out[['2007']])
summary(out[['1952']])

names(out)
```

The iterations do not have to explicitly be over sequential numbers.

```{r, for-nonsequential}
for (yr in years) {
 cat(yr, "\n")
}
```
Nor do you need to iterate over numbers.
```{r nonnumeric}
for (lab in c("bilirubin", "creatinine", "WBC")){
  print(lab)
}

```

If you want to store the result of each loop somewhere, you need to initialize the object.

```{r for_versus_vectorized}
heights_cm <- c(165, 180, 150, 172)
height_in <- c()

for (pt in 1:length(heights_cm)){
  height_converted <- heights_cm[pt] / 2.54
  height_in <- c(height_in, height_converted)
}
print(height_in)
heights_cm /2.54
```
We've done similar calculations using R's vectorized calculation abilities. 

Which option is preferable? We can use the function `system.time` to test.

```{r timing_code}
heights_cm <- sample(140:190, 1000000, replace = T)
heights_in <- c()
# vectorized calculations
system.time(
  heights_in <- heights_cm / 2.54
)

# for loop
system.time(
  for (pt in 1:length(heights_cm)){
    height_converted <- heights_cm[pt] / 2.54
    heights_in <- c(height_in, height_converted)
  }
)

# for loop, initializing the output vector
heights_in <- rep(0, length(heights_cm))
system.time(
  for (pt in 1:length(heights_cm)){
    height_converted <- heights_cm[pt] / 2.54
    heights_in[pt] <- height_converted
  }
)

```

Loops can be slow! We can optimize them to an extent, but they will usually be 
slower than vectorized calculations. However, loops can be very powerful and 
sometimes they are the right approach:

* Need to use a function that doesn't support vectorized operations (and you 
  cannot re-write it)
* The loop option is easier to read than the alternative, and
* Code performance is a low priority or dataset is small

# Branching (if-then-else syntax)

We already saw an example of branching in the gap-minder *for* loop example.

Here's a simple example to illustrate the syntax. 

```{r, if}
val <- rnorm(1)
val
if (val < 0) {
  print("val is negative")
} else {
  print("val is positive")
}
```

We can chain together `if` statements using `else if `.

```{r, if-chain}
val <- rnorm(1)
val
if (val < -1) {
  print("val is more than one standard deviation below the mean.")
} else if (abs(val) <= 1) {
  print("val is within one standard deviation of the mean.")
} else {
  print("val is more than one standard deviation above the mean.")
}
```

In general, the `{` brackets are only needed if you have multiple R expressions,
but R will complain when an `else` starts a line of code, so generally using the
`{` is good practice. That said, this _works just fine_:

```{r, if-oneline}
if (val < 0) print("val is negative") else print("val is positive")
```

# The condition in an if statement

The condition in the if statement cannot be NA or R will give an error. This is
a very common bug.

```{r, if-bug,eval=FALSE}
continents <- unique(gap$continent)
continents

continents <- unique(as.character(gap$continent))
continents <- c('Antarctica', continents)

out <- rep(0, length(continents))
for (i in seq_along(continents)) {
    sub <- filter(gap, continent == continents[i])
    if(mean(sub$lifeExp) < 50)
       print(continents[i])
}

print(i)
sub <- gap[gap$continent == continents[i], ]
if(mean(sub$lifeExp) < 50) print('here')
mean(sub$lifeExp) < 50
```

An `NA`/`NaN` is the main reason an if statement may fail, because R will
generally convert other values to logical values.

Zero evaluates to `FALSE`, all other numbers evaluate to `TRUE`. In general
strings are not converted to booleans.

A more robust alternative is to use `isTRUE()`:

```{r}
isTRUE(NA)
isTRUE(NULL)
isTRUE(1)
isTRUE(FALSE)
isTRUE(TRUE)
isTRUE(as.logical(1))
```

```{r, if-bug-fix}
out <- rep(0, length(continents))
for (i in seq_along(continents)) {
  sub <- filter(gap, continent == continents[i])
  if(isTRUE(mean(sub$lifeExp) > 70)){
    print(continents[i])
  }
}
```


# Vectorized if/else

For "simple" if/else applications, or where vectorized operations need to occur, 
the function `ifelse` may be preferable.

```{r alternative_one_line_if}
print(
  ifelse(val < 0, "val is negative", "val is positive")
)

bilirubin <- data.frame(
  pt = c("M.H.", "P.J.", "K.G."),
  lab = c(1, 1.8, 0.7)
)

bilirubin

bilirubin %>% 
  mutate(
    bilirubin_test = ifelse(bilirubin$lab > 1.2, "elevated", "normal")
  )

bilirubin

```

# Functions

Functions are one of the most important constructs in R (and many other 
languages). They allow you to modularize your code - encapsulating a set of 
repeatable operations as an individual function call.

You should rely heavily on functions rather than having long sets of expressions 
in R scripts.

Functions have many important advantages:

* They reduce bugs by avoiding having multiple instances of the same 
  functionality.
  
* They reduce time involved in coding by eliminating redundancy.

* They make for cleaner and more easily-readable code.

A basic goal is writing functions is *modularity*.

In general, a function should

* be fairly short,

* be focused and specific in what it does, and

* be designed so that it can be used in combination with other functions to 
  carry out more complicated operations.

# Writing functions

## The Basic Syntax

Create a function:

```{r}
repeat_what_i_said <- function(what_i_said){
  paste0("You said: '", what_i_said, "'")
}

repeat_what_i_said("use functions often!")
repeat_what_i_said("cats")
```

A function takes 0 or more arguments, and will return the last unassigned line 
of code, as above, or you can more explicitly return a value.

```{r}

how_much_do_i_like_cats <- function(num){
  if (num < 5) {
    return("You don't like cats very much")
  } else if (num < 50) {
    return("You think cats are okay")
  } else {
    return("You really like cats!")
  }
}

how_much_do_i_like_cats(sample(1:100, 1))

```

## A more complicated example

Let's say that body mass index (BMI) is something you often have to calculate. 
You could remember the formula and type it out each time: bmi = weight/height**2

But it might be less error prone to write a function! 

```{r}
calculate_bmi <- function(height_in_cm, weight_in_kg){
  height_in_m <- height_in_cm/100
  weight_in_kg / height_in_m ** 2
}

calculate_bmi

```

Lets try it out!

```{r}
head(starwars)

starwars %>%
  select(name, height, mass) %>%
  mutate(bmi = calculate_bmi(height_in_cm = height, weight_in_kg = mass))
```

# Function arguments

R can match arguments by name (when provided) or by position (the fall-back). 

```{r}
starwars %>%
  select(name, height, mass) %>%
  mutate(bmi = calculate_bmi(height, mass))
```

It also allows one to specify default values so that the user doesn't have to 
explicitly provide all the arguments.

```{r, fun-args}
calculate_bmi2 <- function(height, weight_in_kg, height_unit = "cm"){
  if (!height_unit %in% c("cm", "m")) {
    stop("Unrecognized unit for height!")
  }
  if (height_unit == "cm") {
    height_in_m <- height/100
  } else {
    height_in_m <- height
  }
  weight_in_kg / height_in_m ** 2
}

calculate_bmi2(170, 70)
calculate_bmi2(1.7, 70, "m")
calculate_bmi2(170, 70, "cm")
calculate_bmi2(67, 70, "inches")
```

# What is the "..." argument for?

Using `...` as one of the arguments to a function allows a function to pass
along user-provided arguments without specifying explicitly what the user might
provide. 

```{r, usedots, fig.cap = ""}
bilirubin_labs <- c(1.0, 1.5, 2.0, NA, 1.8, 0.2, 0.3, 0.7, NA)

mean_above_thresh <- function(vec, thresh, ...){
  vec <- vec[vec > thresh]
  return(mean(vec, ...))
}
?mean
mean_above_thresh(vec = bilirubin_labs, thresh = 1.2, na.rm = T)
```


# Important concepts with R functions

Functions in R return an object. In general, R functions are and should be 
designed such that the only effect of the function is through the return value.

**Side effects** are when a function affects the state of the world in addition 
to its return value.  Can you think of any side effects that you saw an R 
function produce from earlier in the workshop?  What about:

- `library()`
- `setwd()`
- `plot()`

Functions in R are (roughly) *pass-by-value* and not *pass-by-reference*. 
This means that if you modify an argument inside the function it will not change 
the original value outside the function. This protects you from a major 
potential source of side effects. (There are exceptions to this rule.)

In actuality, functions in R are *call-by-value*. What this means for our 
purposes is that you can pass an input argument in without a copy being made of 
it. This saves time and memory. At the time that you modify the input within the 
function (if ever), then a copy is made and  the modified input is different 
than the original value outside the function.

# Variable scope and global variables

In general functions should not make use of variables from outside the function. 
(However, for quick-and-dirty work and in some other circumstances, one may do 
this.) This provides modularity and reduces bugs and surprises.

If R can't find a variable that is used in a function based on the function 
arguments and variables defined locally in the function, it goes and looks 
elsewhere following a set of rules called *lexical scoping*. (This type of 
scoping has to do with R's roots (and explains why R is very similar to other 
languages for functional programming) - we won't go into details here but 
certainly worth looking into as you start using R more.)

Basically this means that it looks for variables relative to where the function 
is defined (not relative to where the function is called).

This can get involved, but a couple brief examples illustrate the basic idea.

```{r, scoping}
x <- 2
f <- function(y) {
    return(x + y)
}
f(1) # 2 + 1 = 3
f(2) # 2 + 2 = 4

# f looks outside the function for something called x, since there is no x 'within' the function

g <- function(y) {
  x <- 10
  return(f(y))
}
g(1) # 2 + 1 = 3
g(2) # 2 + 2 = 4
# f is using the global values for x (f is defined globally)

g <- function(y) {
  f <- function(y) {
     return(x + y)
  }
  x <- 10
  return(f(y))
}

g(1) # 10 + 1 = 11
g(2) # 10 + 2 = 12

#f is defined within g, x is defined within g

```

This is a contrived example - but it's quite common to accidentally use 
variables you didn't define and get unintended errors in your calculations....

Note that `x` is used as a global variable here, which in general is bad 
practice. Don't write functions like `f` :)

# When do I start programming?

> “[W]e wanted users to be able to begin in an interactive environment,
> where they did not consciously think of themselves as programming.
> Then as their needs became clearer and their sophistication increased,
> they should be able **to slide gradually into programming, when the
> language and system aspects would become more important**.”

[John Chambers, quoted by Roger Peng in his UseR Keynote](https://simplystatistics.org/2018/07/12/use-r-keynote-2018/)

# Key Principles of R

- Everything that exists is an object.
- Everything that happens is a function call.

### What does the 2nd principle mean?

Are arithmetic operations really just functions?
```{r, plusfun}
3 + 2
'+'(3, 2)
```

Yes!

And what about indexing?

```{r, indexingfun}
x <- runif(100)
x[2]
'['(x , 2)
```

Also yes!

(But avoid writing your code this way because it isn't super readable.)


### What does the 1st principle mean?

```{r, funs-as-objects}
class(1)
class(runif)
class(function(x) x^2)
square <- function(x) x^2
class(square)
```


### Onwards: Readings and References

* A great reference for learning both basic and advanced concepts in using the R 
language for data analysis is the book _R for Data Science_, by Garrett 
Grolemund and Hadley Wickham. An online version is freely available and may be 
accessed [here](http://r4ds.had.co.nz/). In particular, 
[chapter 21 ("Iteration")](http://r4ds.had.co.nz/iteration.html) is a great 
review of much of what we have covered in this module.


# Breakout

### Basics

1) Write a for loop that will loop through a vector of values and produce a vector equal to the input vector except with negative values set to zero. (ex: `c(-2.3, 0, 6, 9) should become c(0, 0, 6, 9))

```{r}

```

2) Write an R function that will take an input vector and set any negative values in the vector to zero.

```{r }

```

### Using the ideas

2) Write an R function that will take an input vector and set any value below a 
threshold to be the value of threshold. Optionally, the function should instead 
set values above a threshold to the value of the threshold.

```{r}


```

3) Augment your function so that it checks that the input is a numeric vector
and return an error if not. (See the help information for `stop()`.)

```{r}


```
