---
title: 'Module 9: Writing Good Code & Advanced R'
author: "Jasmine Hughes"
date: "9/17/2021"
output: html_document
---

```{r chunksetup, include=FALSE} 
library(dplyr)
gap <- read.csv(
  file.path("..", "data", "gapminder-FiveYearData.csv"), 
  stringsAsFactors = FALSE
) 
```

# Before we start this module...

You know most of what you need to know already!
These rules are just to help you reach your goals with minimal headaches...

Not all of these tips will be immediately useful, but I think the exposure to 
someof these concepts will help you know what tools are at your disposal.

# Style Guides
> Good coding style is like using correct punctuation. You can manage without 
it, but it sure makes things easier to read. - Hadley Wickham

'Advanced R' by Hadley Wickham: http://adv-r.had.co.nz/Style.html

Working groups & companies may have their own!
Google: https://google.github.io/styleguide/Rguide.html

People have strong opinions about this kind of thing:
https://simplystatistics.org/2018/07/27/why-i-indent-my-code-8-spaces/

Overall, Code That Works > Pretty Code... 
... But code is typically read more often than it is written.

# Tips for avoiding bugs

- Plan out your code in advance, including all special cases/possibilities.
  Can you abstract some tasks to make it more general? Can you break down one
  big task into many smaller tasks (and functions)?

- When doing software development, write tests for your code early in
  the process. (There's packages for this! `testthat` is probably the most
  widely used). Tests are essentially "when I give this function x, I expect y".

- Build up code in pieces, testing along the way. Make big changes in
  small steps, sequentially checking to see if the code has broken on
  test case(s). 

- Analysis should also be built up in pieces. Check your data distributions
  regularly. Peak at the first few lines regularly.

- Code in a modular fashion, making good use of functions, so that you
  don't need to debug the same code multiple times. Smaller functions
  are easier to debug, easier to understand, and can be combined in
  a modular fashion.

- Use core R functionality and algorithms already coded. Figure out
  if a functionality already exists in (or can be adapted from) an R
  package (or potentially in a C/Fortran library/package): code that
  is part of standard mathematical/numerical packages will probably
  be more efficient and bug-free than anything you would write.

- Remove objects you don't need, to avoid accidentally using values
  from an old object via the scoping rules.(`remove()`)

- Be careful that the conditions of *if* statements and the sequences
  of *for* loops are robust when they involve evaluating R code.

- Don't hard code numbers - use variables (e.g., number of iterations,
  parameter values in simulations), even if you don't expect to change
  the value, as this makes the code more readable and reduces bugs when
  you use the same number multiple times. 

- Check that inputs to and outputs from functions (either functions
  you call or functions you write) are valid and use `warning()`
  and `stop()` to give a warning or stop execution when something
  unexpected happens.

- Write code for clarity and accuracy first; then worry about efficiency.
  Write an initial version of the code in the simplest way, without
  trying to be efficient; then make a second version that employs efficiency
  tricks and check that both produce the same output.
  
# Common syntax errors and bugs
- Parenthesis mis-matches
  - `[...]` vs. `[[...]]`
  
- Comparing real numbers exactly using `==` is dangerous. 
  Suppose you generate `x = 0.333333` in some fashion with some code and then 
  check: `x == 1/3`. This will produce FALSE. (Try using `round()` to compare 
  numbers if you are in this situation)

- Vectors vs. single values: 
    + `||` vs. `|` and `&&` vs `&`.
    + You expect a single value but your code gives you a vector
    + You want to compare an entire vector but your code just compares the first 
      value (e.g., in an `if` statement) 
    + consider using `identical()` or `all.equal()` or `all()` or `any()` etc.
      
```{r}
c(TRUE,TRUE,FALSE) | c(TRUE)

all(c(TRUE,TRUE,FALSE)) || TRUE

```

- Silent type conversion when you don't want it, or lack of coercion where 
  you're expecting it

- Using the wrong function or variable name

- Giving unnamed arguments to a function in the wrong order (best practices to 
  name them.... easier to read, but less succinct)
  
- Forgetting to define a variable in the environment of a function and having 
  the function, via R's scoping rules, get that variable as a global variable 
  from one of the enclosing environments. 
  
  At best the types are not compatible and you get an error; 
  at worst, you use a garbage value and the bug is hard to trace. 
  In some cases your code may work fine when you develop the code 
  (if the variable exists in the enclosing environment), but then may not work 
  when you restart R if the variable no longer exists or is different.

- NULL vs NA. If a value doesn't exist, and you try to compare it (ex, 
  `is.na(NULL)`), you'll get an error. Try checking first for NULL before NA 
  (`if (is.null(x) | is.na(x))`) or use `isTRUE(is.na(x))`.
  
# Catching bugs

There's a function called `debug` that helps you step through your code to find 
the source of errors. There's a function called `traceback` that tells you which 
functions were called, and in what order.

Let's briefly see these tools in action. I'll demo this in a very basic way, but 
hopefully this will give you an idea of the power of these tools.

```{r eval = F}
buggyFun <- function(myDF) {
   print(names(myDF))
   myDF$id <- seq_len(nrow(myDF))
   sums <- rowSums(myDF)
   return(sums)
}

buggyFun(gap)
# unhelpful error message...

traceback()

debug(buggyFun) #turn on debugger
# run the command below from the console
buggyFun(gap)
undebug(buggyFun) # turn off again

options(error = recover)
buggyFun(gap)

```  

We can also insert ``browser()`` inside a function and R will stop there and 
allow us to proceed with debugging statements.

# Testing

Testing should be performed on multiple levels and begun as early as possible
in the development process.  For programs that accept input either from a user
or file, it is important that the code validates the input is what
it expects to receive. Tests that ensure individual code elements (e.g., 
functions, classes, and class methods) behave correctly are called *unit tests*.

Writing unit tests early in the process of implementing new functionality
helps you think about what you want a piece of code to do, rather than just how
it does it. This practice improves code quality by focusing your attention
on use cases rather than getting lost in implementation details.

The *testthat* package and *testit* package are helpful for setting up tests. 
Also, *RUnit* is a testing framework for R that helps automate test setup, 
creation, execution, and reporting. 


# Memory use

You should know how much memory (RAM) the computer you are using has and keep in 
mind how big your objects are and how much memory you code might use. All 
objects in R are stored in RAM unlike, e.g., SAS or a database.

If in total, the jobs on a machine approach the physical RAM, the machine will 
start to use the hard disk as 'virtual memory'. This is called paging or 
swapping, and once this happens you're often toast (i.e., your code may take 
essentially forever to finish).

Often it's a good idea to roughly estimate how much memory an object will take 
up even before creating it in R. You can do this with some simple arithmetic. 
Every real number takes 8 bytes (integers and logicals take less; character 
strings are complicated), so an object with, say, 1 million rows and 10 columns, 
all numbers, would take roughly 8 * 1000000 * 10 bytes or 800 Mb.

```{r}
x <- rnorm(1e7)
object.size(x)
1e7*8/1e6  # direct calculation of Mb
print(object.size(x), units = 'auto')
x <- rnorm(1e8)
gc()
rm(x)
gc()
```

# Scripting

* Keep your code in script (i.e., text) files.
* Keep your files modular and focused.
* Write functions to reuse code.
* Learn a powerful, general purpose text editor

If you use a good editor (such as RStudio's built-in editor, or Visual Studio 
Code), it's easier to write and understand your code.

With such editors, you can generally then execute lines or blocks of code easily.

To run all the code in an entire file, do `source('myCodeFile.R')`.

# Good coding practices: functions

Use functions whenever possible. In particular try to write functions 
rather than carry out your work using blocks of code. Why? Functions 
allow us to reuse blocks of code easily for later use and for recreating 
an analysis (reproducible research). It's more transparent than sourcing a file
of code because the inputs and outputs are specified formally, so
you don't have to read through the code to figure out what it does.

Good use of functions includes:

- Write reusable code for core functionality and keep a single copy
  of the code (w/ backups of course) so you only need to change it in one place

- Smaller functions are easier to debug, easier to understand, and can
  be combined in a modular fashion (like the UNIX utilities). 
  (Goal: 50 or fewer lines)

Functions should: 

 - be modular (having a single task); 
 
 - have meaningful name; and
 
 - have a comment describing their purpose, inputs and outputs (see the
   help file for any standard R function for how this is done in that context).
   See `roxygen2` for the package that makes these look pretty. 
   https://cran.r-project.org/web/packages/roxygen2/vignettes/roxygen2.html

Two options:
1. Include your functions in your analysis scripts up at the top.
2. Create them in a new .R file, add them to environment with 
`source('pathtofile/functionfile.R')`

The first is easiest, but once you build up a code base, the second is preferred. 
(And makes it easier to compile functions into a package...)

# Good coding practices: syntax

- Header information: put meta info on the code into the first few lines
  of the file as comments. Include who, when, what, how the code fits
  within a larger program (if appropriate), possibly the versions of
  R and key packages that you wrote this for

- Indentation: do this systematically (your editor can help here). This
  helps you and others to read and understand the code and can help
  in detecting errors in your code because it can expose lack of symmetry.

- Whitespace: use a lot of it. Some places where it is good to have
  it are (1) around operators (assignment and arithmetic), (2) between
  function arguments and list elements, (3) between matrix/array indices,
  in particular for missing indices. 

- Use blank lines to separate blocks of code and comments to say what
  the block does

- Split long lines at meaningful places. (80 character length is traditional)

```{r}
ggplot(gap) + 
  aes(x = factor(year), y = lifeExp) + 
  geom_boxplot()
```
  
- Use parentheses for clarity even if not needed for order of operations. 
  For example, `a/y*x` will work but is not easy to read and
  you can easily induce a bug if you forget the order of operations.
  
- Documentation - add lots of comments (but don't belabor the obvious). 
  Remember that in a few months, you may not follow 
  your own code any better than a stranger. Some key things to document: 
    - summarizing a block of code, 
    - explaining a very complicated piece of code 
    - explaining arbitrary constant values.
    - break code into separate files (< 1000 lines per file) with meaningful 
      file names and related functions grouped within a file.
- Choose a consistent naming style for objects and functions: 
  e.g. *numIts* (lowerCamelCase) vs. *NumIts* (UpperCamelCase) 
  vs. *num.its* vs. *num\_its*
    + I'd suggest avoiding periods in names since periods are used for 
      object-oriented programming in R and many other languages
      
    - Try to have the names be informative without being overly long.
    
    - Try to avoid using the names of standard R functions for your objects, 
      but R will generally be fairly smart about things.
      
```{r}
c <- 7
c(3,5)
c
rm(c)
c
```

- Use active names for functions (e.g., *calc_loglik*, *calcLogLik*)


# Reproducible research

> An article about computational science in a scientific
publication is **not the scholarship itself**, it is merely
**advertising** of the scholarship. The actual scholarship is the
complete software development environment and the
complete set of instructions which generated the figures.
-  Jonathan Buckheit and David Donoho, WaveLab and Reproducible Research (1995)

Here are some useful articles talking about reproducibility.

- [Wilson et at., Best practices for scientific computing, ArXiv:1210:0530](http://arxiv.org/abs/1210.0530)
- [Gentzkow and Shapiro tutorial for social scientists](https://web.stanford.edu/~gentzkow/research/CodeAndData.pdf)


# Some ideas for improving reproducibility

- Never change a dataset manually, including in pre-processing and 
  post-processing. 
  * Always have a script that operates on the data (or results/output). 
  * It's a good idea to have your projects organized into folders, with one 
    folder for your raw data, and one folder for your "cleaned" or "processed" 
    data, whatever that might mean.

- Produce figures (e.g., from R) via a script and not by point-and-click.

- When making figures, use `save()` or `save.image()` to save all the inputs 
  needed to recreate a figure, with the code for making the figure in a script 
  file.
- If feasible, include your code for doing analyses and making figures in the 
  relevant document reporting the work by using one of the following tools
    - *R Markdown*
    - *Latex* with *knitr* 
    - *Jupyter* (also works with python and julia)
    
- Always set the random number seed so someone else can duplicate your exact 
  numbers.
  
- Use version control tools such as Git! 

# Some good books

Many are available online for free!
 
R in general:

* Wickham, Hadley; Advanced R. [https://adv-r.had.co.nz](https://adv-r.had.co.nz) 
  (good coverage of advanced topics in the R language)
  
* Chambers, John; Software for Data Analysis: Programming with R (More 
  conceptual/theoretical)

Specific aspects of R:

* Wilke, Claus; Fundamentals of Data Visualization. https://serialmentor.com/dataviz/
* Xie, Yihui; Dynamic documents with R and knitr. https://github.com/yihui/knitr-book
* Murrell, Paul; R Graphics, 2nd ed. [https://www.stat.auckland.ac.nz/\~paul/RG2e](https://www.stat.auckland.ac.nz/\~paul/RG2e/)
* Murrell, Paul; Introduction to Data Technologies. [https://www.stat.auckland.ac.nz/\~paul/ItDT/](https://www.stat.auckland.ac.nz/\~paul/ItDT/)
* Wickham, Hadley; R Packages. http://r-pkgs.had.co.nz/ For building your own packages.
